

<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{rentrez tutorial}
-->
 
```{r, eval=TRUE, echo=FALSE}
library(knitr)
opts_knit$set(upload.fun = image_uri)
opts_knit$set(fig.keep = "last")
opts_knit$set(fig.show = "hold")

```
 

Rentrez tutorial
================

rentrez, entrez and the NCBI
----------------------------

`rentrez` allows you to acess the [NCBI's Entrez Utilities API](http://www.ncbi.nlm.nih.gov/books/NBK25500/) 
, and thus the tens of millions of records in databases like GenBank and Pubmed,
from  the comfort of an R session. Specifically, this packages provies functions
that search the NCBI's databases, discover cross-referenced records, and fetch 
records from the NCBI. 


Getting started with the rentrez
--------------------------------

`rentrez` gives you a few tools that help you to get to know your way around the
NCBI's databases. First, you can use `entrez_dbs()` to find out just what
databases are avaliable:

```{r, dbs}
library(rentrez)
entrez_dbs()
```

You can find out a bit more about each of those database using a set of
functions with names starting `entrez_db_`:

| Function name            | Return                                               |
|--------------------------|------------------------------------------------------|
| `entrez_db_summary()`    | Brief description of what the database is            |
| `entrez_db_searchable()` | Set of search terms that can used with this database |
| `entrez_db_links() `     | Set of databases that might contain linked records   |

For instance, we can find out a little more about the cryptically named
database 'cdd'...

```{r, cdd}
entrez_db_summary("cdd")
```

... or fin
d out which search terms can be used with the Short Read Archive (SRA)
sequencing database:

```{r, sra_eg}
entrez_db_searchable("sra")
```

Searching databases: `entrez_search`
------------------------------------

The function  `entrez_search()` to discover records in each available database.
You provide a database name (`db`) and a search term(`term`). If you print the
returned object you get a summary of what it contains, and the way the NCBI's
servers translated your query:

```{r eg_search}
r_search <- entrez_search(db="pubmed", term="R Language")
r_search
```
There are a few things to note here. First, the NCBI's server has worked out
that we meant R as a programming language, and so inlcuded the 
['MeSh' term](http://www.ncbi.nlm.nih.gov/mesh). We'll worry about these and
other search terms in a second. Second, there are many more 'hits' for this search than there
are unique IDs contained in this object. That's because the optional argument
`retmax`, which controls the maximum number of returned values has a default
value of 20. We can acess those 20 IDs like this:


```{r search_ids}
r_search$ids
```

And get more IDs by increasing the `ret_max` argument. 

```{r searchids_2}
another_r_search <- entrez_search(db="pubmed", term="R Language", retmax=40)
another_r_search
```

If we want to get IDs for all of those thousands of records that match this
search, we can use the NCBI's web history feature (DESCRIBED IN LINK). When
the argument `use_history` to `'y'` the object returned by entrez_search
contains `






```{r

