---
title: Rentrez Tutorial
author: "David winter"
date: "`r Sys.Date()`"
output: 
    rmarkdown::html_vignette:
        toc: true
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---
```{r, count_recs, echo=FALSE}
library(rentrez)
count_recs <- function(db, denom) {
    nrecs <-  rentrez::entrez_db_summary(db)["Count"]
    round(as.integer(nrecs)/denom, 1)
}
```
## Introduction: The NCBI, entrez and `rentrez`.

The NCBI shares a _lot_ of data. At the time this document was compiled, there
were `r count_recs("pubmed",1e6)` million papers in [PubMed](http://www.ncbi.nlm.nih.gov/pubmed/),
including `r count_recs("pmc", 1e6)` million full-text records available in [PubMed Central](http://www.ncbi.nlm.nih.gov/pubmed/).
[The NCBI Nucleotide Database](http://www.ncbi.nlm.nih.gov/nuccore) (which includes GenBank) has data for `r count_recs("nuccore", 1e6)`
million different sequences, and [dbSNP](http://www.ncbi.nlm.nih.gov/snp/) describes
`r count_recs("snp", 1e6)` million different genetic variants. All of these
records can be cross-referenced with the  `r round(entrez_search(db="taxonomy", term='species[RANK]')$count/1e6,2)` million
species in the [NCBI taxonomy](www.ncbi.nlm.nih.gov/taxonomy) or `r count_recs("omim", 1e3)` thousand disease-associated records
in [OMIM](http://www.ncbi.nlm.nih.gov/omim).


The NCBI makes this data available through a [web interface](http://www.ncbi.nlm.nih.gov/),
an [FTP server](ftp://ftp.ncbi.nlm.nih.gov/) and through a REST API called the
[Entrez Utilities](http://www.ncbi.nlm.nih.gov/books/NBK25500/) (`Eutils` for
short). This package provides functions to use that API, allowing users to
gather and combine data from multiple NCBI databases in the comfort of an R
session or script.

## Getting started with the rentrez

To make the most of all the data the NCBI shares you need to know a little about
their databases, the records they contain and the ways you can find those
records. The [NCBI provides extensive documentation for each of their
databases](http://www.ncbi.nlm.nih.gov/home/documentation.shtml) and for the
[EUtils API that `rentrez` takes advantage of](http://www.ncbi.nlm.nih.gov/books/NBK25501/).
There are also some helper functions in `rentrez` that help users learn their
way around the NCBI's databases. 

First, you can use `entrez_dbs()` to find the list of available databases:

```{r, dbs}
entrez_dbs()
```
There is a set of functions with names starting `entrez_db_` that can be used to
gather more information about each of these databases:

**Functions that help you learn about NCBI databases**

| Function name            | Return                                               |
|--------------------------|------------------------------------------------------|
| `entrez_db_summary()`    | Brief description of what the database is            |
| `entrez_db_searchable()` | Set of search terms that can used with this database |
| `entrez_db_links() `     | Set of databases that might contain linked records   |

For instance, we can get a description of the somewhat cryptically named
database 'cdd'...

```{r, cdd}
entrez_db_summary("cdd")
```

... or find out which search terms can be used with the Sequence Read Archive (SRA)
database (which contains raw data from sequencing projects):

```{r, sra_eg}
entrez_db_searchable("sra")
```

Just how these 'helper' functions might be useful will become clearer once
you've started using `rentrez`, so let's get started.

## Searching databases: `entrez_search()`

Very often, the first thing you'll want to do with `rentrez` is search a given
NCBI database to find records that match some keywords. You can do this using
the function `entrez_search()`. In the simplest case you just need to provide a 
database name (`db`) and a search term (`term`) so let's search PubMed for
articles about the `R language`:


```{r eg_search}
r_search <- entrez_search(db="pubmed", term="R Language")
```
The object returned by a search acts like a list, and you can get a summary of 
its contents by printing it.

```{r print_search}
r_search
```

There are a few things to note here. First, the NCBI's server has worked out
that we meant R as a programming language, and so included the
['MeSH' term](http://www.ncbi.nlm.nih.gov/mesh) term associated with programming
languages. We'll worry about MeSH terms and other special queries later, for now
just note that you can use this feature to check that your search term was interpreted in the way
you intended. Second, there are many more 'hits' for this search than there
are unique IDs contained in this object. That's because the optional argument
`retmax`, which controls the maximum number of returned values has a default
value of 20.

The IDs are the most important thing returned here. They
allow us to fetch records matching those IDs, gather summary data about them or find
cross-referenced records in other databases. We access the IDs as a vector using the
`$` operator:


```{r search_ids}
r_search$ids
```

If we want to get more than 20 IDs we can do so by increasing the `ret_max` argument.

```{r searchids_2}
another_r_search <- entrez_search(db="pubmed", term="R Language", retmax=40)
another_r_search
```

If we want to get IDs for all of the thousands of records that match this
search, we can use the NCBI's web history feature [described below](#web_history).


### Building search terms

The EUtils API uses a special syntax to build search terms. You can search a
database against a specific term using the format `query[SEARCH FIELD]`, and
combine multiple such searches using the boolean operators `AND`, `OR` and `NOT`.

For instance, we can find next generation sequence datasets for the (amazing...) ciliate
_Tetrahymena thermophila_ by using the organism ('ORGN') search field:


```{r, Tt}
entrez_search(db="sra",
              term="Tetrahymena thermophila[ORGN]",
              retmax=0)
```

We can narrow our focus to only those records that have been added recently (using the colon to 
specify a range of values):


```{r, Tt2}
entrez_search(db="sra",
              term="Tetrahymena thermophila[ORGN] AND 2013:2015[PDAT]",
              retmax=0)
```

Or include recent records for either _T. thermophila_ or it's close relative _T.
borealis_ (using parentheses to make ANDs and ORs explicit).


```{r, Tt3}
entrez_search(db="sra",
              term="(Tetrahymena thermophila[ORGN] OR Tetrahymena borealis[ORGN]) AND 2013:2015[PDAT]",
              retmax=0)
```

The set of search terms available varies between databases. You can get a list
of available terms or any given data base with `entrez_db_searchable()`

```{r, sra_searchable}
entrez_db_searchable("sra")
```

###Precise queries using MeSH terms

In addition to the search terms described above, the NCBI allows searches using
[Medical Subject Heading (MeSH)](http://www.ncbi.nlm.nih.gov/mesh) terms. These
terms create a 'controlled vocabulary',  and allow users to make very finely
controlled queries of databases.

For instance, if you were interested in reviewing studies on how a class of
anti-malarial drugs called Folic Acid Antagonists work against _Plasmodium vivax_ (a
particular species of malarial parasite), you could use this search:

```{r, mesh}
entrez_search(db   = "pubmed",
              term = "(vivax malaria[MeSH]) AND (folic acid antagonists[MeSH])")
```

The complete set of MeSH terms is available as a database from the NCBI. That
means it is possible to download detailed information about each term and find 
the ways in which terms relate to each other using `rentrez`. You can search 
for specific terms with `entrez_search(db="mesh", term =...)` and learn about the
results of your search using the tools described below.

### Advanced counting

As you can see above, the  object returned by `entrez_search()` includes the
number of records matching a given search. This means you can learn a little
about the composition of, or trends in, the records stored in the NCBI's
databases using only the search utility. For instance, let's track the rise of
the scientific buzzword "connectome" in PubMed, programmatically creating 
search terms for the `PDAT` field:

```{r, connectome, fig.width=5, fig.height=4, fig.align='center'}
search_year <- function(year, term){
    query <- paste(term, "AND (", year, "[PDAT])")
    entrez_search(db="pubmed", term=query, retmax=0)$count
}

year <- 2008:2014
papers <- sapply(year, search_year, term="Connectome", USE.NAMES=FALSE)

plot(year, papers, type='b', main="The Rise of the Connectome")
```

## Finding cross-references : `entrez_link()`:


One of the strengths of the NCBI databases is the degree to which records of one
type are connected to  other records within the NCBI or to external data
sources. The function `entrez_link()` allows users to discover these links
between records.

###My god, it's full of links

To get an idea of the degree to which records in the NCBI are cross-linked we
can find all NCBI data associated with a single gene (in this case the 
Amyloid Beta Precursor gene, the product of which is associated with the 
plaques that form in the brains of  Alzheimer's Disease patients).

The function `entrez_link()` can be used to find cross-referenced records. In
the most basic case we need to provide an ID (`id`), the database from which this
ID comes (`dbfrom`) and the name of a database in which to find linked records (`db`).
If we set this last argument to 'all' we can find links in multiple databases:

```{r elink0}
all_the_links <- entrez_link(dbfrom='gene', id=351, db='all')
all_the_links
```
Just as with `entrez_search` the returned object behaves like a list, and we can
learn a little about its contents by printing it. In the case, all of the
information is in `links` (and there's a lot of them!):


```{r elink_link}
all_the_links$links
```
The names of the list elements are in the format `[source_database]_[linked_database]` 
and the elements themselves contain a vector of linked-IDs. So, if we want to
find open access publications associated with this gene we could get linked records
in PubMed Central:

```{r, elink_pmc}
all_the_links$links$gene_pmc[1:10]
```

Or if were interested in this gene's role in diseases we could find links to OMIM:

```{r, elink_omim}
all_the_links$links$gene_omim

```

###Narrowing our focus

If we know beforehand what sort of links we'd like to find , we can
to use the `db` argument to narrow the focus of a call to `entrez_link`.

For instance, say we are interested in knowing about all of the 
RNA transcripts associated with the Amyloid Beta Precursor gene in humans. 
Transcript sequences are stored in the nucleotide database (referred
to as `nuccore` in EUtils), so to find transcripts associated with a given gene
we need to set `dbfrom=gene` and `db=nuccore`.

```{r, elink1}
nuc_links <- entrez_link(dbfrom='gene', id=351, db='nuccore')
nuc_links
nuc_links$links
```
The object we get back contains links to the nucleotide database generally, but
also to special subsets of that database like [refseq](http://www.ncbi.nlm.nih.gov/refseq/). 
We can take advantage of this narrower set of links to find IDs that match unique
transcripts from our gene of interest.

```{r, elinik_refseqs}
nuc_links$links$gene_nuccore_refseqrna
```
We can use these ids in calls to `entrez_fetch()` or `entrez_summary()` to learn
more about the transcripts they represent. 

###External links

In addition to finding data within the NCBI, `entrez_link` can turn up
connections to external databases. Perhaps the most interesting example is
finding links to the full text of papers in PubMed. For example, when I wrote
this document the first paper linked to Amyloid Beta Precursor  had a unique ID of
`25500142`. We can find links to the full text of that paper with `entrez_link` 
by setting the `cmd` argument to 'llinks':

```{r, outlinks}
paper_links <- entrez_link(dbfrom="pubmed", id=25500142, cmd="llinks")
paper_links
```

Each element of the `linkouts` object contains information about an external
source of data on this paper:

```{r, urls}
paper_links$linkouts
```

Each of those linkout objects contains quite a lot of information, but the URL 
is probably the most useful. For that reason, `rentrez` provides the
function `linkout_urls` to make extracting just the URL simple:

```{r just_urls}
linkout_urls(paper_links)
```

The full list of options for the `cmd` argument are given in in-line
documentation (`?entrez_link`). If you are interested in finding full text
records for a large number of articles checkout the package 
[fulltext](https://github.com/ropensci/fulltext) which makes use of multiple
sources (including the NCBI) to discover the full text articles.

###Using more than one ID

It is possible to pass more than one ID to `entrez_link()`. By default, doing so
will give you a single elink object containing the complete set of links for 
_all_ of the IDs that you specified. So, if you were looking for protein IDs
related to specific genes you could do:

```{r, multi_default}
all_links_together  <- entrez_link(db="protein", dbfrom="gene", id=c("93100", "223646"))
all_links_together
all_links_together$links$gene_protein
```

Although this behaviour might sometimes be useful, it means we've lost track of
which `protein` ID is linked to which `gene` ID. To retain that information we
can set `by_id` to `TRUE`. This gives us a list of elink objects, each once
containing links from a single `gene` ID:

```{r, multi_byid}
all_links_sep  <- entrez_link(db="protein", dbfrom="gene", id=c("93100", "223646"), by_id=TRUE)
all_links_sep
lapply(all_links_sep, function(x) x$links$gene_protein)
```


## Getting summary data: `entrez_summary()`

Having found the unique IDs for some records via `entrez_search` or `entrez_link()`, you are
probably going to want to learn something about them. The `Eutils` API has two
ways to get information about a record. `entrez_fetch()` returns 'full' records
in varying formats and `entrez_summary()` returns less information about each
record, but in relatively simple format. Very often the summary records have the information
you are after, so `rentrez` provides functions to parse and summarise summary
records.


###The summary record

`entrez_summary()` takes a vector of unique IDs for the samples you want to get
summary information from. Let's start by finding out something about the paper
describing [Taxize](https://github.com/ropensci/taxize), using its PubMed ID:


```{r, Summ_1}
taxize_summ <- entrez_summary(db="pubmed", id=24555091)
taxize_summ
```

Once again, the object returned by `entrez_summary` behaves like a list, so you can extract
elements using `$`. For instance, we could convert our PubMed ID to another
article identifier...

```{r, Summ_2}
taxize_summ$articleids
```
...or see how many times the article has been cited in PubMed Central papers

```{r, Summ_3}
taxize_summ$pmcrefcount
```

###Dealing with many records

If you give `entrez_summary()` a vector with more than one ID you'll get a
list of summary records back. Let's get those _Plasmodium vivax_ papers we found
in the `entrez_search()` section back, and fetch some summary data on each paper:

```{r, multi_summ}
vivax_search <- entrez_search(db = "pubmed",
                              term = "(vivax malaria[MeSH]) AND (folic acid antagonists[MeSH])")
multi_summs <- entrez_summary(db="pubmed", id=vivax_search$ids)
```

`rentrez` provides a helper function, `extract_from_esummary()` that takes one
or more elements from every summary record in one of these lists. Here it is
working with one...

```{r, multi_summ2}
extract_from_esummary(multi_summs, "fulljournalname")
```
... and several elements:

```{r, multi_summ3}
date_and_cite <- extract_from_esummary(multi_summs, c("pubdate", "pmcrefcount",  "title"))
knitr::kable(head(t(date_and_cite)), row.names=FALSE)
```

##Fetching full records: `entrez_fetch()`

As useful as the summary records are, sometimes they just don't have the
information that you need. If you want a complete representation of a record you
can use `entrez_fetch`, using the argument `rettype` to specify the format you'd
like the record in.

###Fetch DNA sequences in fasta format

Let's extend the example given in the `entrez_link()` section about finding
transcript for a given gene. This time we will fetch cDNA sequences of those
transcripts.We can start by repeating the steps in the earlier example
to get nucleotide IDs for refseq transcripts of two genes:

```{r, transcript_ids}
gene_ids <- c(351, 11647)
linked_seq_ids <- entrez_link(dbfrom="gene", id=gene_ids, db="nuccore")
linked_transripts <- linked_seq_ids$links$gene_nuccore_refseqrna
head(linked_transripts)
```

Now we can get our sequences with `entrez_fetch`, setting `rettype` to "fasta"
(the list of formats available for [each database is give in this table](http://www.ncbi.nlm.nih.gov/books/NBK25499/table/chapter4.T._valid_values_of__retmode_and/)):

```{r fetch_fasta}
all_recs <- entrez_fetch(db="nuccore", id=linked_transripts, rettype="fasta")
class(all_recs)
nchar(all_recs)
```

Congratulations, now you have a really huge character vector! Rather than
printing all those thousands of bases we can take a peak at the top of the file:

```{r, peak}
cat(strwrap(substr(all_recs, 1, 500)), sep="\n")
```

If we wanted to use these sequences in some other application we could write them 
to file:

```r
write(all_recs, file="my_transcripts.fasta")
```

Alternatively, if you want to use them  within an R session  
we could write them to a  temporary file then read that. In this case I'm using `read.dna()` from the
pylogenetics package ape (but not executing the code block in this vignette, so
you don't have to install that package):

```r
temp <- tempfile()
write(all_recs, temp)
parsed_recs <- ape::read.dna(all_recs, temp)
```

###Fetch a parsed XML document

Most of the NCBI's databases can return records in XML format. In additional to
downloading the text-representation of these files, `entrez_fetch()` can return
objects parsed by the `XML` package. As an example, we can check out the  Taxonomy
database's record for (did I mention they are amazing....) _Tetrahymena
thermophila_, specifying we want the result to be parsed by setting
`parsed=TRUE`:

```{r, Tt_tax}
Tt <- entrez_search(db="taxonomy", term="(Tetrahymena thermophila[ORGN]) AND Species[RANK]")
tax_rec <- entrez_fetch(db="taxonomy", id=Tt$ids, rettype="xml", parsed=TRUE)
class(tax_rec)
```

The package XML (which you have if you have installed `rentrez`) provides
functions to get information from these files. For relatively simple records
like this one you can use `XML::xmlToList`:

```{r, Tt_list}
tax_list <- XML::xmlToList(tax_rec)
tax_list$Taxon$GeneticCode
```

For more complex records, which generate deeply-nested lists, you can use
[XPath expressions](https://en.wikipedia.org/wiki/XPath) along with the function 
`XML::xpathSApply` or the extraction operatord `[` and `[[` to extract specific parts of the
file. For instance, we can get the scientific name of each taxon in _T.
thermophila_'s lineage by specifying a path through the XML

```{r, Tt_path}
tt_lineage <- tax_rec["//LineageEx/Taxon/ScientificName"]
tt_lineage[1:4]
```

As the name suggests, `XML::xpathSApply()` is a counterpart of base R's
`sapply`, and can be used to apply a function to
nodes in an XML object. A particularly useful function to apply is `XML::xmlValue`, 
which returns the content of the node:

```{r, Tt_apply}
XML::xpathSApply(tax_rec, "//LineageEx/Taxon/ScientificName", XML::xmlValue)
```
There are a few more complex examples of using `XPath` [on the rentrez wiki](https://github.com/ropensci/rentrez/wiki)

<a name="web_history"></a>

##Using NCBI's Web History features

When you are dealing with very large queries it can be time  consuming to pass
long vectors of unique IDs to and from the NCBI. To avoid this problem, the NCBI
provides a feature called "web history" which allows users to store IDs on the
NCBI servers then refer to them in future calls.

###Post a set of IDs to the NCBI for later use: `entrez_post()`

If you have a list of many NCBI IDs that you want to use later on, you can post
them to the NCBI's severs. In order to provide a brief example, I'm going to post just one
ID, the `omim` identifier for asthma:

```{r, asthma}
upload <- entrez_post(db="omim", id=600807)
upload
```
The NCBI sends you back some information you can use to refer to the posted IDs. 
In `rentrez`, that information is represented as a `web_history` object. 

###Get a `web_history` object from `entrez_search` or `entrez_link()`

In addition to directly uploading IDs to the NCBI, you can use the web history
features with `entrez_search` and `entrez_link`. For instance, imagine you wanted to
find all of the sequences of the widely-studied gene COI from all snails
(which are members of the taxonomic group Gastropoda):

```{r, snail_search}
entrez_search(db="nuccore", term="COI[Gene] AND Gastropoda[ORGN]")
```

That's a lot of sequences! If you really wanted to download all of these it
would be a good idea to save all those IDs to the server by setting
`use_history` to `TRUE` (note you now get a `web_history` object along with your
normal search result):

```{r, snail_history}
snail_coi <- entrez_search(db="nuccore", term="COI[Gene] AND Gastropoda[ORGN]", use_history=TRUE)
snail_coi
snail_coi$web_history
```

Similarity, `entrez_link()` can return `web_history` objects by using the `cmd`
`neighbor_history`. Let's find SNPs associated with asthma (using the `omim` ID
mentioned above):

```{r, asthma_links}
asthma_snps <- entrez_link(dbfrom="omim", db="snp", cmd="neighbor_history", id=600807)
asthma_snps$web_histories
```

As you can see, instead of returning lists of IDs for each linked database (as
it would be default), `entrez_link()` now  returns a list of web_histories.

###Use a `web_history` object

Once you have those IDs stored on the NCBI's servers, you are going to want to
do something with them. The functions `entrez_fetch()` `entrez_summary()` and
`entrez_link()` can all use `web_history` objects in exactly the same way they
use IDs. 

So, we could repeat the last example (finding SNPs linked to asthma), but this
time using the ID we uploaded earlier

```{r, asthma_links_upload}
asthma_snps <- entrez_link(dbfrom="omim", db="snp", cmd="neighbor_history", web_history=upload)
asthma_snps
```

... then summarize each linked SNP using `entrez_summary()`:
                           

```{r, links}
snp_summ <- entrez_summary(db="snp", web_history=asthma_snps$web_histories$omim_snp)
knitr::kable(extract_from_esummary(snp_summ, c("chr", "fxn_class", "global_maf")))
```

If you really wanted to you could also use `web_history` objects to download all those thousands of COI sequences.
When downloading large sets of data, it is a good idea to take advantage of the
arguments `retmax` and `restart` to split the request up into smaller chunks.
For instance, we could get the first 200 sequences in 50-sequence chunks:

(note: this code block is not executed as part of the vignette to save time and bandwidth):


```r
for( seq_start in seq(1,200,50)){
    recs <- entrez_fetch(db="nuccore", web_history=snail_coi$web_history,
                         rettype="fasta", retmax=50, retstart=seq_start)
    cat(recs, file="snail_coi.fasta", append=TRUE)
    cat(seq_start+49, "sequences downloaded\r")
}
```

##What next ?

This tutorial has introduced you to the core functions of `rentrez`, there are
almost limitless ways that you could put them together. [Check out the wiki](https://github.com/ropensci/rentrez/wiki)
for more specific examples, and be sure to read the inline-documentation for
each function. If you run into problem with rentrez, or just need help with the
package and `Eutils` please contact us by opening an issue at the [github
repository](https://github.com/ropensci/rentrez/issues)
 


